---
title: "390 EDA Project"
author: "Fahmi Khalid"
date: "April 25, 2017"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(stringr)
employee <- read_csv("WA_Fn-UseC_-HR-Employee-Attrition.csv")
quit <- filter(employee, Attrition=="Yes")
stayed <- filter(employee, Attrition=="No")


gender_influence <- employee %>% group_by(Gender, JobRole) %>% summarise(average_monthly_income=mean(MonthlyIncome))
gender_influence_monthlyincome <- gender_influence %>% mutate(round_income=round(average_monthly_income))

ggplot(gender_influence_monthlyincome, aes(x=Gender,y=average_monthly_income, color=Gender))+
    geom_bar(stat = "identity",width=0.5,position=position_dodge())+
    geom_text(aes(label=round_income), vjust=1.6, color="white",
              position = position_dodge(0.9), size=3.5)+
    facet_wrap(~JobRole)+
    theme_minimal()+
    scale_fill_manual(values=c("red", "blue"))+
    ggtitle("Monthly incomes across Job Roles")+
    scale_y_continuous(name="Average Monthly Income")

```

Graph of average monthly income across all roles faceted by gender.

```{r cars}

Job_roles <- unique(quit$JobRole)
roles_quit <- str_count(Job_roles)
loss_by_job <- data_frame(Job=Job_roles, "Percentage loss"=(roles_quit/1470)*100)
loss_by_job <- loss_by_job %>% mutate(round_loss=round(loss_by_job$`Percentage loss`,digits=2))
loss_by_job <- arrange(loss_by_job, desc(round_loss))
ggplot(loss_by_job, aes(x=reorder(loss_by_job$Job,-loss_by_job$round_loss),y=loss_by_job$round_loss))+
    geom_bar(stat = "identity",width=0.5, position=position_dodge())+
    geom_text(aes(label=loss_by_job$round_loss), vjust=1.6, color="white",
              position = position_dodge(0.9), size=3.5)+
    theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))+
    labs(x="Job Roles",y="Percentage of Loss")+
  ggtitle("Loss percentage across different Job Roles")


```
Graph of percent of employee lost across job roles.

```{r pressure, echo=FALSE}
testing <- employee %>% group_by(MaritalStatus) %>% summarise(mean=round(mean(MonthlyIncome),digits=0))
ggplot(testing, aes(x=testing$MaritalStatus,y=testing$mean))+
  geom_bar(stat = "identity",width=0.5, position=position_dodge())+
  geom_text(aes(label=testing$mean), vjust=1.6, color="white",
            position = position_dodge(0.9), size=3.5)+
  labs(x="Marital Status",y="Average Monthly Income")+
  ggtitle("Put a ring on it")

```

Average monthly income based on Marital status.



```{r}

RelSat <-employee %>% group_by(RelationshipSatisfaction) %>% summarise(mean=mean(MonthlyIncome))
ggplot(RelSat, aes(x=reorder(RelSat$RelationshipSatisfaction,-RelSat$mean),y=RelSat$mean))+
  geom_bar(stat = "identity",width=0.5, position=position_dodge())+
  geom_text(aes(label=round(RelSat$mean)), vjust=1.6, color="white",
            position = position_dodge(0.9), size=3.5)+
  labs(x="Relationship Satisfaction",y="Average Monthly Income")+
  ggtitle("99 Problems")

```
Average monthly income across different levels of Relationship satisfaction.

```{r}

JobSat <- employee %>% group_by( JobSatisfaction) %>% summarise(mean=mean(MonthlyIncome))
Jobsatty <- c("Low", "Medium", "High", "Very High")
ggplot(JobSat, aes(x=reorder(Jobsatty,-JobSat$mean),y=JobSat$mean))+
  geom_bar(stat = "identity",width=0.5, position=position_dodge())+
  geom_text(aes(label=round(JobSat$mean)), vjust=1.6, color="white",
            position = position_dodge(0.9), size=3.5)+
  labs(x="Job Satisfaction",y="Average Monthly Income")+
  ggtitle("Mo Money Mo Problems")


```

Average monthly income across different levels of Job Satisfaction.

```{r}
lin_reg <- lm(WorkLifeBalance~ YearsSinceLastPromotion, employee)
summary(lin_reg)
```
There is no significant linear relationship between Work life balance and years since last promotion 
```{r}
boxplot(MonthlyIncome~WorkLifeBalance,data = employee,xlab="PerformanceRating", ylab="MonthlyIncome")
```

```{r}
numeric <- employee[, c(1,4,6,7,10)]
library("Hmisc")
#see correlation between features 
res2 <- rcorr(as.matrix(numeric)) # data must be a matrix 

flattenCorrMatrix <- function(cormat, pmat) {
    ut <- upper.tri(cormat)
    data.frame(
        row = rownames(cormat)[row(cormat)[ut]],
        column = rownames(cormat)[col(cormat)[ut]],
        cor  =(cormat)[ut],
        p = pmat[ut]
    )
}
flattenCorrMatrix(res2$r, res2$P)


```
The figure above shows the correlation between 5 features and the significance level. 

```{r}
library("PerformanceAnalytics")
chart.Correlation(numeric, histogram=TRUE, pch=19)
```
The figure above is also the visualization of correlations between 5 features. 
The distribution of each variable is shown on the diagonal.
On the bottom of the diagonal : the bivariate scatter plots with a fitted line are isplayed
On the top of the diagonal : the value of the correlation plus the significance level as stars. 
Each significance level is associated to a symbol : p-values(0, 0.001, 0.01, 0.05, 0.1, 1) <=> symbols(“***”, “**”, “*”, “.”, " “)

```{r}
genderIncome <- employee %>% group_by(Gender) %>%
    summarise(meanSalary = mean(MonthlyIncome))
ggplot (genderIncome, aes (x = Gender, y = meanSalary)) + geom_bar(stat ="identity") + ggtitle("Mean monthly income between gender ")
```
The figure above shows female has higher avarage salary than males in IBM.

```{r}

ggplot(employee, aes(x = Gender, color = Gender)) +
    geom_bar()+ facet_wrap(~ JobRole)+ggtitle("Job distribution among gender")+ ggtitle("Job distribution across gender ")
```

```{r}
genderJobIncome <- employee  %>% group_by(Gender, JobRole) %>%
    summarise(meanSalarybyJob = mean(MonthlyIncome))
ggplot(genderJobIncome, aes(x = JobRole,y = meanSalarybyJob, color = Gender)) +
    geom_point() +ggtitle("Mean monthly income among diffrent jobs across gender ")
```
The figure above illastrates the job distribution between male and female.